Transport map accelerated adaptive importance sampling, and application to inverse problems arising from multiscale stochastic reaction networks

### 2019-01-10 TETAIS.tex
New file to reflect new title
Also changed AMR -> MT throughout

If the ensemble is large enough, and the chain has entered
probabilistic stationarity, then the current state of the ensemble is
a good rough discrete approximation of the target density, and in turn
$\chi^{(k)}$ is close enough to $\pi$ to produce an efficient importance
sample $\{(\hat{\theta}^{(k)},w^{(k)})\}$,  where $\theta^{(k)} =
[\theta_1^{(k)}, \ldots, \theta_M^{(k)}]^\top$ and $w^{(k)} =
[w_1^{(k)}, \ldots, w_M^{(k)}]^\top$. It can be advantageous to use stratified sampling of the
mixture in order to
ensure that the sample made is as representative as possible of the
target density, i.e.
\[ \theta_i^{(k)} \sim q(\cdot ; \theta_i^{(k)},\beta)\]
for each $i = 1,2,\ldots,M$. Usually the kernel $q(\cdot ;
x,\beta)$ is centred at, or has mean equal to $x$, and whose variance
is scaled with $\beta$. We now have a weighted sample, and it
would be inadvisable to use an equally weighted mixture of proposal
distributions from each of these points. Therefore, before starting
the next iteration, the importance sample
$\{(\theta^{(k)},w^{(k)})\}$ is resampled to produce an equally
weighted sample, ready for the next iteration of the algorithm. In
ETAIS, a state-of-the-art resampler is used, which uses optimal
transport methods to find the equally weighted discrete sample which
best represents the statistics of the importance
sample\cite{reich2013nonparametric}. For larger ensemble sizes $M$ this can become
expensive, in which case a greedy approximation of this algorithm, the
Approximate Multinomial Resampler (AMR) can be
implemented\cite{cotter2015parallel}. The output of the resampler is then denoted
$\theta^{(k+1)}$ and the algorithm is ready for the next
iteration. The ETAIS is summarised in Algorithm \ref{alg:ETAIS}, where
the importance samples $\{(\hat{\theta}^{(k)},w^{(k)})\}$ are stored as
the output.

@@

\begin{figure}[!ht]
\centering
\subfigure[Comparison of the two Transport ETAIS
options.]{\includegraphics[width=0.45\textwidth]{images/TPAIS/R1_L2}}
\subfigure[Comparison of Transport ETAIS option (3) with the standard algorithms.]{\includegraphics[width=0.45\textwidth]{images/TPAIS/R1_L2_all}}
\caption{Convergence of algorithms; Transport M-H,
  algorithm \ref{alg:TransportETAIS1}, algorithm \ref{alg:TransportETAIS2} for
  density \eqref{eqn:R2_repeat}. Ensemble size $M=150$, resampling performed using the AMR algorithm.}
\label{fig:R2_l2_convergence}
\end{figure}

@@

% choice of resampler?
We will use the AMR\cite{russ2017parallel} resampler with an ensemble size of
$M=500$. This increase in ensemble size in comparison with previous
example in Section \ref{sec:conv} is to allow for the increase in parameter dimension.

@@

In all eight algorithms we adapt the proposal variances during the
burn-in phase of the algorithms to avoid the need for a huge ensemble size to approximate the posterior with a mixture of isotropic Gaussian kernels.
We use the AMR resampler, this time with an ensemble size of $M=2500$. The increase in ensemble size compensates for the increase in dimension from four to eight.

@@

In one
dimension the ETPF algorithm can be implemented very efficiently. As
described in~\cite{reich2013nonparametric}, the coupling matrix has
all non-zero entries in a staircase pattern when the state space is
ordered. We can exploit this knowledge to produce
Algorithm~\ref{alg:ETPF_1d}, which is much faster than using the
simplex algorithm to minimise the associated cost function, and faster
than the AMR algorithm\cite{cotter2015parallel}. Moreover, the task of
running this resampler can then be parallelised over $d$
processors. This could be an area for further study.

>>>>

If the ensemble is large enough, and the chain has entered
probabilistic stationarity, then the current state of the ensemble is
a good rough discrete approximation of the target density, and in turn
$\chi^{(k)}$ is close enough to $\pi$ to produce an efficient importance
sample $\{(\hat{\theta}^{(k)},w^{(k)})\}$,  where $\theta^{(k)} =
[\theta_1^{(k)}, \ldots, \theta_M^{(k)}]^\top$ and $w^{(k)} =
[w_1^{(k)}, \ldots, w_M^{(k)}]^\top$. It can be advantageous to use stratified sampling of the
mixture in order to
ensure that the sample made is as representative as possible of the
target density, i.e.
\[ \theta_i^{(k)} \sim q(\cdot ; \theta_i^{(k)},\beta)\]
for each $i = 1,2,\ldots,M$. Usually the kernel $q(\cdot ;
x,\beta)$ is centred at, or has mean equal to $x$, and whose variance
is scaled with $\beta$. We now have a weighted sample, and it
would be inadvisable to use an equally weighted mixture of proposal
distributions from each of these points. Therefore, before starting
the next iteration, the importance sample
$\{(\theta^{(k)},w^{(k)})\}$ is resampled to produce an equally
weighted sample, ready for the next iteration of the algorithm. In
ETAIS, a state-of-the-art resampler is used, which uses optimal
transport methods to find the equally weighted discrete sample which
best represents the statistics of the importance
sample\cite{reich2013nonparametric}. For larger ensemble sizes $M$ this can become
expensive, in which case a greedy approximation of this algorithm, the
Multinomial Transformation (MT) can be
implemented\cite{cotter2015parallel}. The output of the resampler is then denoted
$\theta^{(k+1)}$ and the algorithm is ready for the next
iteration. The ETAIS is summarised in Algorithm \ref{alg:ETAIS}, where
the importance samples $\{(\hat{\theta}^{(k)},w^{(k)})\}$ are stored as
the output.

@@

\begin{figure}[!ht]
\centering
\subfigure[Comparison of the two Transport ETAIS
options.]{\includegraphics[width=0.45\textwidth]{images/TPAIS/R1_L2}}
\subfigure[Comparison of Transport ETAIS option (3) with the standard algorithms.]{\includegraphics[width=0.45\textwidth]{images/TPAIS/R1_L2_all}}
\caption{Convergence of algorithms; Transport M-H,
  algorithm \ref{alg:TransportETAIS1}, algorithm \ref{alg:TransportETAIS2} for
  density \eqref{eqn:R2_repeat}. Ensemble size $M=150$, resampling performed using the MT algorithm.}
\label{fig:R2_l2_convergence}
\end{figure}

@@

% choice of resampler?
We will use the MT\cite{russ2017parallel} resampler with an ensemble size of
$M=500$. This increase in ensemble size in comparison with previous
example in Section \ref{sec:conv} is to allow for the increase in parameter dimension.

@@

In all eight algorithms we adapt the proposal variances during the
burn-in phase of the algorithms to avoid the need for a huge ensemble size to approximate the posterior with a mixture of isotropic Gaussian kernels.
We use the MT resampler, this time with an ensemble size of $M=2500$. The increase in ensemble size compensates for the increase in dimension from four to eight.

@@

In one
dimension the ETPF algorithm can be implemented very efficiently. As
described in~\cite{reich2013nonparametric}, the coupling matrix has
all non-zero entries in a staircase pattern when the state space is
ordered. We can exploit this knowledge to produce
Algorithm~\ref{alg:ETPF_1d}, which is much faster than using the
simplex algorithm to minimise the associated cost function, and faster
than the MT algorithm\cite{cotter2015parallel}. Moreover, the task of
running this resampler can then be parallelised over $d$
processors. This could be an area for further study.
